---
title: "Data Science Crash Course 4 Practitioners"
subtitle: "Session 2 - Intro to Data Processing"
author: "Aaron Scherf"
date: "July 19, 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Today's Commands



## 0. Intro to Data Processing

There is a common adage among data scientists that at least 70% of your time in the program will be spent cleaning and processing data.

Thus it is important to learn the functions and workflow of processing data early so you can become experienced (and therefore faster) at it.

The typical steps in data processing can be summarized as follows, as explained in [this article from Towards Data Science](https://towardsdatascience.com/the-ultimate-guide-to-data-cleaning-3969843991d4):

- Inspect Data Quality
- Clean Data to Fix Anomalies
- Verify Data Quality
- Report on Changes to Dataset

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

